---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# wcvpmatch

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![CRAN status](https://www.r-pkg.org/badges/version/wcvpmatch)](https://CRAN.R-project.org/package=wcvpmatch)
[![](http://cranlogs.r-pkg.org/badges/grand-total/wcvpmatch?color=green)](https://cran.r-project.org/package=wcvpmatch)
[![](http://cranlogs.r-pkg.org/badges/last-week/wcvpmatch?color=green)](https://cran.r-project.org/package=wcvpmatch)
[![R-CMD-check](https://github.com/PaulESantos/wcvpmatch/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PaulESantos/wcvpmatch/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`wcvpmatch` is an R package for scientific plant name standardization and
taxonomic reconciliation against the [World Checklist of Vascular Plants (WCVP)](https://powo.science.kew.org/about-wcvp).

The package is inspired by the matching workflow implemented in
`treemendous` (especially its staged `matching` logic), and extends that
approach with broader functionality for robust and reproducible taxonomic
resolution workflows.

## What `wcvpmatch` adds

- robust parsing and normalization of scientific names (`classify_spnames()`)
- staged exact + fuzzy matching at genus/species/infraspecies levels (`wcvp_matching()`)
- infraspecific rank-aware reconciliation (trinomial support)
- optional genus prefiltering for performance
- duplicate-aware matching with full row traceability (`allow_duplicates`, `input_index`)
- taxonomic context output (`matched_taxon_name`, `accepted_taxon_name`, status fields)
- author metadata propagation from WCVP (`matched_taxon_authors`, `accepted_taxon_authors`)
- optional standardized output names in snake_case (`output_name_style = "snake_case"`)

## Installation

`wcvpmatch` depends on `fozziejoin` for fuzzy matching. At the moment, `fozziejoin` is installed from GitHub and requires a working Rust toolchain.

### 1) Install Rust

Windows (GNU toolchain required for R + Rtools compatibility):

```bash
rustup install stable-x86_64-pc-windows-gnu
rustup default stable-x86_64-pc-windows-gnu
rustup target add x86_64-pc-windows-gnu --toolchain stable-x86_64-pc-windows-gnu
```

Linux / macOS: install Rust with the official installer at <https://www.rust-lang.org/tools/install>.

### 2) Install the development version of `fozziejoin` from GitHub:

```r
pak::pak("fozzieverse/fozziejoin/fozziejoin-r")
```

Additional information is available in the official [`fozziejoin`](https://github.com/fozzieverse/fozziejoin/tree/main/fozziejoin-r) GitHub repository.


### 3) Install the development version of `wcvpdata` from GitHub: 

```r
pak::pak("PaulESantos/wcvpdata")
```

### 4) Install the development version of `wcvpmatch` from GitHub: 

```r
pak::pak("PaulESantos/wcvpmatch")
```

## Quick start

```{r}
library(wcvpmatch)

sp_names <- c(
  "Aniba heterotepala",          
  "Anthurium quipuscoae",        
  "Centropogon reflexus",        
  "Chuquiraga jonhstonii",       
  "Cyathea carolinae",           
  "Ditassa violascens",   
  "Cleistocactus fieldianus",
  "Opuntia yanganucensis",
  "Epidendrum trachydipterum",  
  "Hebeclinium hylophorbum",     
  "Jaltometa sagastegui",        # Jaltomata sagastegui
  "Lepechinia tomentosa",        
  "Lupinus cookianos",           # Lupinus cookianus
  "Oxalis hochreutinerii",       # Oxalis hochreutineri
  "Passiflora heterohelix",     
  "Peperomia arborigaudens",     # Peperromia arborigaudens
  "Piper setulosum",             
  "Pycnophyllum aristattum",     # Pycnophyllum aristatum
  "Salvia subscadens",           # Salvia subscandens
  "Stellaria macbridei",         
  "Stemodia piurenses",          # Stemodia piurensis
  "Weberbauerella brongnartioides" 
)


res <- 
  classify_spnames(sp_names) |> 
  wcvp_matching(
  prefilter_genus = TRUE,
  allow_duplicates = TRUE,
  max_dist = 2,
  method = "osa",
  output_name_style = "snake_case"
)

res |>  dim()

res |> 
  dplyr::select(input_name,
                matched_taxon_name,
                accepted_taxon_name, 
                taxon_status) |> 
  as.data.frame()

```

## Matching workflow

`wcvp_matching()` follows a staged pipeline:

1. `wcvp_direct_match()`
2. `wcvp_genus_match()`
3. `wcvp_fuzzy_match_genus()`
4. `wcvp_direct_match_species_within_genus()`
5. `wcvp_suffix_match_species_within_genus()`
6. `wcvp_fuzzy_match_species_within_genus()`
7. infraspecific checks (rank + fuzzy infraspecies, when applicable)

This staged design prioritizes strict evidence first, then progressively
relaxes criteria to recover additional valid matches.


## Input requirements

`wcvp_matching()` accepts:

- parsed output from `classify_spnames()`
- or a tibble/data.frame with minimal columns `Genus` and `Species`

Optional columns:

- `Infra.Rank`, `Infraspecies` for trinomial matching
- `Input.Name` to preserve original query text
- `Author` (from parser) as contextual information in output

## Output highlights

`wcvp_matching()` returns:

- traceable rows via `input_index`
- original taxon components (`Orig.*`) and matched components (`Matched.*`)
- pathway flags (`direct_match`, `fuzzy_match_genus`, etc.)
- final logical indicator `matched`
- taxonomic context fields:
- `matched_plant_name_id`
- `matched_taxon_name`
- `matched_taxon_authors`
- `taxon_status`
- `accepted_plant_name_id`
- `accepted_taxon_name`
- `accepted_taxon_authors`
- `is_accepted_name`

## Example: FIA data and ambiguous genus ties

When input genera have tied fuzzy candidates, `wcvp_matching()` emits:

`Multiple fuzzy matches for some genera (tied distances). The first match is selected.`

Example with an FIA-like table (`Orig.Genus`, `Orig.Species`):

```{r}
fia

fia_result <- fia |>
  wcvp_matching(
    allow_duplicates = TRUE,
    max_dist = 2
  ) |>
  dplyr::select(
    input_name,
    orig_genus,
    matched_genus,
    orig_species,
    matched_species,
    taxon_status,
    accepted_taxon_name
  )
fia_result
```

Inspect ambiguous genus cases from the result attribute:

```{r}
amb_g <- attr(fia_result, "ambiguous_genus")
amb_g
```

Re-run only those ambiguous names for focused review:

```{r}
amb_g |>
  dplyr::ungroup() |>
  dplyr::select(Orig.Genus, Orig.Species) |>
  wcvp_matching(
    allow_duplicates = TRUE,
    max_dist = 2
  ) |>
  dplyr::select(
    input_name,
    orig_genus,
    matched_genus,
    orig_species,
    matched_species,
    taxon_status,
    accepted_taxon_name
  )
```

In this example, the ambiguous genus spellings (`Howeia`, `Hyeronima`) are
resolved to accepted genera (`Howea`, `Hieronyma`) after species-level checks.


## Performance notes

For medium/large inputs, recommended settings are:

```r
wcvp_matching(
  parsed_df,
  allow_duplicates = TRUE
)
```

This typically reduces runtime by limiting candidate genera before species-level
matching and avoids recomputing duplicated names.



## Core public functions

- `classify_spnames()`
- `wcvp_matching()`
- `build_genus_index()`
- `prefilter_target_by_genus()`
- `wcvp_direct_match()`
- `wcvp_genus_match()`
- `wcvp_fuzzy_match_genus()`
- `wcvp_direct_match_species_within_genus()`
- `wcvp_fuzzy_match_species_within_genus()`
- `wcvp_suffix_match_species_within_genus()`

## Acknowledgement

`wcvpmatch` builds on ideas used in the [`treemendous`](https://github.com/speckerf/treemendous) matching workflow and
extends them for WCVP-focused reconciliation, richer parser diagnostics, and
reproducible row-level traceability.
