---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)

if (file.exists("DESCRIPTION") && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else {
  library(wcvpmatch)
}

library(tibble)
library(dplyr)
```

# wcvpmatch

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![CRAN status](https://www.r-pkg.org/badges/version/wcvpmatch)](https://CRAN.R-project.org/package=wcvpmatch)
[![](http://cranlogs.r-pkg.org/badges/grand-total/wcvpmatch?color=green)](https://cran.r-project.org/package=wcvpmatch)
[![](http://cranlogs.r-pkg.org/badges/last-week/wcvpmatch?color=green)](https://cran.r-project.org/package=wcvpmatch)
<!-- badges: end -->

`wcvpmatch` standardizes scientific plant names and reconciles them against a
World Checklist of Vascular Plants (WCVP)-style backbone. It combines parsing,
exact and fuzzy matching, accepted-name resolution, and optional distribution
retrieval in a single workflow.

The package is built around three main pieces:

- `classify_spnames()` parses and normalizes submitted names.
- `wcvp_matching()` resolves names against a WCVP-like backbone.
- `wcvp_distribution()` retrieves species, genus, or family distribution from
  WCVP name and distribution tables.

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("PaulESantos/wcvpmatch")
```

Install the CRAN release:

```r
# install.packages("pak")
pak::pak("wcvpmatch")
```

`wcvpmatch` uses `fozziejoin` for fuzzy matching. If `fozziejoin` is installed
from source, a working Rust toolchain is needed.

Install Rust from <https://rust-lang.org/tools/install/>. On Windows, the most
practical setup for R + Rtools is:

```bash
rustup override set stable-x86_64-pc-windows-gnu
```

Install `fozziejoin`:

```r
pak::pak("fozziejoin")
```

To use the default WCVP backbone automatically, install `wcvpdata` from
`r-universe`:

```r
install.packages(
  "wcvpdata",
  repos = c("https://paulesantos.r-universe.dev", "https://cloud.r-project.org")
)
```

## Quick example: `wcvp_matching()`

```{r}
matching_backbone <- tibble(
  genus = c("Aniba", "Jaltomata", "Veronica", "Veronica"),
  species = c("heterotepala", "sagastegui", "vulcanica", "spathulata"),
  infraspecific_rank = NA_character_,
  infraspecies = NA_character_,
  plant_name_id = c(1, 2, 10, 200),
  taxon_name = c(
    "Aniba heterotepala",
    "Jaltomata sagastegui",
    "Veronica vulcanica",
    "Veronica spathulata"
  ),
  taxon_authors = c("A.Author", "B.Author", "C.Author", "D.Author"),
  taxon_status = c("Accepted", "Accepted", "Synonym", "Accepted"),
  accepted_plant_name_id = c(1, 2, 200, 200)
)

matching_result <- classify_spnames(
  c("Aniba heterotepala", "Jaltometa sagasteguii", "Veronica vulcanica")
) |>
  wcvp_matching(
    target_df = matching_backbone,
    allow_duplicates = TRUE,
    max_dist = 2,
    method = "osa",
    add_name_distance = TRUE,
    output_name_style = "snake_case"
  ) |>
  select(
    input_name,
    matched_taxon_name,
    accepted_taxon_name,
    taxon_status,
    matched_dist
  )

matching_result
```

## Quick example: `wcvp_distribution()`

```{r}
distribution_names <- tibble(
  plant_name_id = c(1, 2, 3, 4, 5, 6),
  accepted_plant_name_id = c(NA, 3, NA, NA, 1, NA),
  taxon_rank = c("Species", "Species", "Species", "Species", "Species", "Species"),
  taxon_status = c("Accepted", "Synonym", "Accepted", "Accepted", "Synonym", "Accepted"),
  family = c("Cactaceae", "Cactaceae", "Cactaceae", "Fagaceae", "Cactaceae", "Cactaceae"),
  genus = c("Opuntia", "Nopalea", "Opuntia", "Quercus", "Opuntia", "Mammillaria"),
  species = c("ficus-indica", "cochenillifera", "cochenillifera", "robur", "tuna", "elongata"),
  taxon_name = c(
    "Opuntia ficus-indica",
    "Nopalea cochenillifera",
    "Opuntia cochenillifera",
    "Quercus robur",
    "Opuntia tuna",
    "Mammillaria elongata"
  )
)

distribution_records <- tibble(
  plant_locality_id = 1:7,
  plant_name_id = c(1, 2, 3, 3, 4, 5, 6),
  continent_code_l1 = c("8", "8", "8", "4", "1", "8", "8"),
  continent = c(
    "SOUTHERN AMERICA", "SOUTHERN AMERICA", "SOUTHERN AMERICA",
    "NORTHERN AMERICA", "EUROPE", "SOUTHERN AMERICA", "SOUTHERN AMERICA"
  ),
  region_code_l2 = c("83", "83", "83", "41", "10", "85", "83"),
  region = c(
    "Western South America", "Western South America", "Western South America",
    "Mexico", "Europe", "Southern South America", "Western South America"
  ),
  area_code_l3 = c("MEX", "PER", "COL", "MEX", "ESP", "GAL", "MEX"),
  area = c("Mexico", "Peru", "Colombia", "Mexico", "Spain", "Galapagos", "Mexico"),
  introduced = c(0, 0, 0, 1, 0, 0, 0),
  extinct = c(0, 0, 0, 0, 0, 0, 0),
  location_doubtful = c(0, 0, 0, 0, 0, 0, 0)
)

distribution_result <- wcvp_distribution(
  c("Nopalea cochenilliferaa", "Taxon inexistente"),
  taxon_rank = "species",
  summarise_by_input = TRUE,
  wcvp_names = distribution_names,
  wcvp_distributions = distribution_records
) |>
  select(
    submited_name,
    accepted_taxon_name,
    distribution_status,
    distribution,
    n_areas
  )

distribution_result
```

## Learn more

The `README` keeps the examples short. For full guides, see the package
vignettes:

- `vignette("wcvp-matching", package = "wcvpmatch")`
- `vignette("wcvp-distribution", package = "wcvpmatch")`

Those articles describe:

- accepted-name resolution and status handling
- staged fuzzy matching and diagnostics
- duplicate handling and profiling
- species, genus, and family distribution retrieval
- occurrence filters and summarised output

## Acknowledgement

`wcvpmatch` builds on ideas used in the
[`treemendous`](https://github.com/speckerf/treemendous) matching workflow and
extends them for WCVP-focused reconciliation and reproducible row-level
traceability.
