---
title: "Retrieve Distribution with wcvp_distribution()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieve Distribution with wcvp_distribution()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

if (file.exists("../DESCRIPTION") && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all("..", export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else {
  library(wcvpmatch)
}

library(tibble)
library(dplyr)
```

`wcvp_distribution()` retrieves geographic distribution from a pair of WCVP-like
tables: one names table and one distribution table. In regular use those tables
can come from `wcvpdata::wcvp_matching_names()` and
`wcvpdata::wcvp_distribution()`, but the function also works with
user-supplied tables that follow the same schema.

It always returns a plain tibble, never an `sf` object. The row-level result
retains `area_code_l3`, the WGSrpd level-3 identifier to use when geometries are
joined in a separate spatial workflow.

This vignette uses a compact reproducible example so every code block runs
without depending on external data packages.

## Example backbone

```{r}
make_distribution_names <- function() {
  tibble(
    plant_name_id = c(1, 2, 3, 4, 5, 6),
    accepted_plant_name_id = c(NA, 3, NA, NA, 1, NA),
    taxon_rank = c("Species", "Species", "Species", "Species", "Species", "Species"),
    taxon_status = c("Accepted", "Synonym", "Accepted", "Accepted", "Synonym", "Accepted"),
    family = c("Cactaceae", "Cactaceae", "Cactaceae", "Fagaceae", "Cactaceae", "Cactaceae"),
    genus = c("Opuntia", "Nopalea", "Opuntia", "Quercus", "Opuntia", "Mammillaria"),
    species = c("ficus-indica", "cochenillifera", "cochenillifera", "robur", "tuna", "elongata"),
    taxon_name = c(
      "Opuntia ficus-indica",
      "Nopalea cochenillifera",
      "Opuntia cochenillifera",
      "Quercus robur",
      "Opuntia tuna",
      "Mammillaria elongata"
    )
  )
}

make_distribution_records <- function() {
  tibble(
    plant_locality_id = 1:7,
    plant_name_id = c(1, 2, 3, 3, 4, 5, 6),
    continent_code_l1 = c("8", "8", "8", "4", "1", "8", "8"),
    continent = c(
      "SOUTHERN AMERICA", "SOUTHERN AMERICA", "SOUTHERN AMERICA",
      "NORTHERN AMERICA", "EUROPE", "SOUTHERN AMERICA", "SOUTHERN AMERICA"
    ),
    region_code_l2 = c("83", "83", "83", "41", "10", "85", "83"),
    region = c(
      "Western South America", "Western South America", "Western South America",
      "Mexico", "Europe", "Southern South America", "Western South America"
    ),
    area_code_l3 = c("MEX", "PER", "COL", "MEX", "ESP", "GAL", "MEX"),
    area = c("Mexico", "Peru", "Colombia", "Mexico", "Spain", "Galapagos", "Mexico"),
    introduced = c(0, 0, 0, 1, 0, 0, 0),
    extinct = c(0, 0, 0, 0, 0, 0, 0),
    location_doubtful = c(0, 0, 0, 0, 0, 0, 0)
  )
}

distribution_names <- make_distribution_names()
distribution_records <- make_distribution_records()

distribution_names
distribution_records
```

## Species-level retrieval

At species level, `wcvp_distribution()` parses input names, matches them through
the same backend used by `wcvp_matching()`, resolves accepted names when
possible, and then joins the result to the distribution table.

```{r}
species_out <- wcvp_distribution(
  c("Nopalea cochenilliferaa", "Taxon inexistente"),
  taxon_rank = "species",
  wcvp_names = distribution_names,
  wcvp_distributions = distribution_records
)

species_out |>
  select(
    submited_name,
    matched_taxon,
    accepted_taxon_name,
    area_code_l3,
    area,
    distribution_status
  )
```

The first query is fuzzy-matched to a synonym and then resolved to the accepted
taxon `Opuntia cochenillifera`. The second query is preserved in the output with
`distribution_status = "no_match"`.

## One row per submitted name

When `summarise_by_input = TRUE`, the function collapses the area-related
columns to one row per input taxon.

```{r}
species_summary <- wcvp_distribution(
  c("Nopalea cochenilliferaa", "Taxon inexistente"),
  taxon_rank = "species",
  summarise_by_input = TRUE,
  wcvp_names = distribution_names,
  wcvp_distributions = distribution_records
)

species_summary |>
  select(
    submited_name,
    accepted_taxon_name,
    distribution_status,
    area_codes,
    distribution,
    n_areas
  )
```

This format is useful for reporting and export because `distribution`,
`areas`, `regions`, and `continents` are returned as collapsed text values
separated by `" - "`.

## Compact table for a spatial join

Use `output = "spatial"` to retain one row per taxon-area combination while
removing repeated matching and geographic-label columns. The resulting
`area_code_l3` remains ready to join to a level-3 polygon layer later.

```{r}
spatial_out <- wcvp_distribution(
  "Nopalea cochenillifera",
  taxon_rank = "species",
  output = "spatial",
  wcvp_names = distribution_names,
  wcvp_distributions = distribution_records
)

spatial_out
```

## Genus-level retrieval

At genus level the function aggregates all accepted taxa in the genus and then
summarises occurrences by area.

```{r}
genus_out <- wcvp_distribution(
  "Opuntia",
  taxon_rank = "genus",
  introduced = FALSE,
  wcvp_names = distribution_names,
  wcvp_distributions = distribution_records
)

genus_out |>
  select(matched_taxon, area_code_l3, area, occurrence_type, distribution_status)
```

The introduced Mexico record is excluded here because `introduced = FALSE`.

## Family-level retrieval

Family-level lookup can also use fuzzy matching through `fozziejoin`.

```{r}
family_out <- wcvp_distribution(
  "Cactacee",
  taxon_rank = "family",
  max_dist = 1,
  wcvp_names = distribution_names,
  wcvp_distributions = distribution_records
)

family_out |>
  select(matched_taxon, match_distance, area_code_l3, area) |>
  distinct()
```

## Fallback from species to genus

If a species query cannot recover species-level distribution, the function can
fall back to the matched genus without interrupting execution.

```{r}
fallback_out <- wcvp_distribution(
  "Opuntia especieinventada",
  taxon_rank = "species",
  wcvp_names = distribution_names,
  wcvp_distributions = distribution_records
)

fallback_out |>
  select(submited_name, matched_taxon, area_code_l3, area, distribution_status)
```

This fallback is marked with `distribution_status = "genus_distribution_fallback"`.

## Practical notes

- Use `summarise_by_input = TRUE` when you need one row per submitted name.
- Prefer `output = "summary"` in new code; `summarise_by_input = TRUE` remains
  available for compatibility.
- Use `output = "spatial"` before joining to a spatial layer; use `output =
  "full"` when matching diagnostics and complete geographic labels are needed.
- Keep the default output when you need one row per taxon-area combination.
- Join a level-3 spatial table later with `left_join(..., by = c("LEVEL3_COD" =
  "area_code_l3"))`; this package intentionally does not create geometries.
- `taxon_rank = "order"` and `taxon_rank = "higher"` are available when the
  names table includes matching `order` and `higher` columns.
- The columns `native`, `introduced`, `extinct`, and `location_doubtful` can be
  used as filters before the final summary is returned.
- If `wcvpdata` is installed, you can omit `wcvp_names` and
  `wcvp_distributions` to use the default backbone.
